Symbolic linearization of equations of motion of constrained multibody systems

### 2013-08-01 PetersonGedeHubbard2012.tex
Incorporated Mont's sect 3 changes.

\section{Dynamics Example}
To illustrate Kane's method in the presence of configuration and velocity
constraints, we consider a thin disc $C$ rolling without slip on a horizontal
plane. The dextral unit vectors $\hat{c}_x, \hat{c}_y, \hat{c}_z$ are fixed to $C$
with $\hat{c}_y$ perpendicular to the disc plane. We take $r$ and $m$ as the
disc radius and mass, and assume a thin disc mass distribution so that
$I_{xx}=I_{zz}=mr^2/4$ and $I_{yy} = mr^2/2$. We take the inertial (Newtonian)
frame $N$ to be similarly equipped with dextral unit vectors $\hat{n}_x,
\hat{n}_y, \hat{n}_z$, with $\hat{n}_z$ perpendicular to the ground plane
(downwards) and aligned with local gravitational field. To orient the disc, we
first align the $C$ with $N$, then apply successive body fixed ZXY rotations:
yaw $q_1$, roll $q_2$, and spin $q_3$.  We denote by $A$ and $B$ the two frames
associated with the first two intermediate rotations in the sequence; we refer
to these as the instantaneous yaw and lean frames, respectively. The unit
vector $\hat{b}_z$ fixed in the $B$ frame is parallel to the disc plane and
aligned with the position vector from the disc center $C^*$ to the lowest point
of the disc $P^*$. Traditionally, to locate the center of the disc $C^*$
relative to the inertial origin $N^*$, two coordinates are used to locate the
contact point in the ground plane; those two coordinates plus the lean angle
$q_2$ determine the location of the center of the disc. The advantage of this
approach is that the choice of coordinates guarantees the disc remain in
contact with the ground plane. To illustrate how dependent coordinates must be
accounted for when linearizing the equations of motion, we purposefully locate
the center of the disc $C^*$ relative to the inertial origin $N^*$ with a
non-minimal choice of coordinates
\begin{equation*}
  \bar{r}^{N^*C^*} = q_4 \hat{n}_x + q_5 \hat{n}_y + q_6 \hat{n}_z
\end{equation*}
which results in the configuration constraint $\mathbf{f}_c$
\begin{equation}
  \label{rd:f_c}
  r\cos{q_2} + q_6 = 0
\end{equation}
which must be satisfied for the disc to remain in contact with the ground.
While this constraint is easily avoided by appropriate choice of coordinates,
it serves to illustrate considerations which must be made when analyzing
systems where there is no obvious way to eliminate configuration constraints.

Again to demonstrate how velocity constraints must be accounted for when
obtaining linearized equations of motion, we define the velocity of $C^*$ in
$N$ with a non-minimal choice of speeds, defined in the body-fixed ($C$) frame.
The angular velocity of $C$ in $N$, however, is defined in the intermediate
frame $B$, which is the instantaneous lean frame.
\begin{align}
  \label{v_u}
  {^N}\bar{v}^{C^*} &= u_4 \hat{c}_x + u_5 \hat{c}_y + u_6 \hat{c}_z \\
  \label{w_u}
  {^N}\bar{\omega}^C &= u_1 \hat{b}_x + u_2 \hat{b}_y + u_3 \hat{b}_z
\end{align}
It should be immediately obvious from these defintions that the partial angular
velocity and partial velocity of $C^*$, respectively, are
\begin{align}
  {^N}\bar{v}^{C^*}_u &= [0, 0, 0, \hat{c}_x, \hat{c}_y, \hat{c}_z] \\
  {^N}\bar{\omega}^C_u &= [\hat{b}_x, \hat{b}_y, \hat{b}_z, 0, 0, 0] 
\end{align}
where we present the partial velocities as a $1\times6$ matrix whose entries
are populated with standard 3-vectors. Given these definitions, the velocity of
the lowest point of the disc $P^*$ is
\begin{align*}
    {^N}\bm{v}^{P^*} &= \bm{v}^{C^*} + \bm{\omega}^{C} \times r^{P^*/C^*} \\
                     &= u_4\hat{c}_x + u_5\hat{c}_y + u_6\hat{c}_z + r u_2
                        \hat{b}_x - r u_1 \hat{b}_y \\
                     &= (r u_2 \cos(q_3) + u_4) \hat{c}_x + (-r
                        u_1 + u_5)\hat{c}_y + r u_2 \sin(q_3) +
                        u6) \hat{c}_z
\end{align*}
where we have made use of the fact that $\bm{r}^{P^*/C^*} = r\bm{b}_z$.
Under the assumption of pure rolling this immediately yields three velocity
constraint equations which make up $\mathbf{f}_v = \mathbf{0}$
\begin{subequations}
\label{rd:f_v}
\begin{align}
    r u_2 \cos(q_3) + u_4 &= 0\\
            -r u_1 + u_5 &= 0\\
    r u_2 \sin(q_3) + u_6 &= 0
\end{align}
\end{subequations}
These equations are linear in the $u_i$ terms, nonlinear in the $q_i$ terms,
involve geometric system parameters (but not mass or inertial parameters), and
do not explicitly involve $\dot{q}_i$ terms. This structure will be taken
advantage of subsequently.

At this point, the reader familiar with the rolling disc problem is certainly
wondering why we have chosen such a diabolical set of coordinates and speeds.
In this simple example, there is an {\bf \textit{obvious and minimal}} choice of
coordinates and speeds ($q_6, u_4, u_5, u_6$, should be dependent; this choice
will be valid for all parameters and configurations; further, the derivation of
the equations of motion can be done without ever introducing these quantities),
for some systems this is not the case (e.g., Stewart's platform, bicycle). The
numeric values of system parameters and the configurations in which the system
will operate will determine which coordinates and speeds can be taken to be
dependent (and hence potentially eliminated by algebraic manipulations).
However, it may not be possible \textit{apriori} to easily determine which
coordinates and speeds should be taken as dependent; further, some systems may
operate in distinct enough regions of the configuration space that one
\textit{must} use different choices of dependent coordinates and/or speeds as
the system moves from one region to another. The purpose of our diabolical
choice of coordinates and speeds is to illustrate, in the context of a familiar
example, how constraints must be taken into account when linearizing equations
of motion. It is our hope that by doing this for a well known and relatively
simple example, readers can apply the same techniques to more complicated
systems where its use is more appropriate or necessary.

>>>>

\section{Dynamics Example}
To illustrate Kane's method in the presence of configuration and velocity
constraints, we consider a thin disc $C$ rolling without slip on a horizontal
plane. The dextral unit vectors $\hat{c}_x, \hat{c}_y, \hat{c}_z$ are fixed to $C$
with $\hat{c}_y$ perpendicular to the disc plane. Let $r$ and $m$ be the
disc radius and mass, and assume a thin disc mass distribution so that
$I_{xx}=I_{zz}=mr^2/4$ and $I_{yy} = mr^2/2$. Take the inertial (Newtonian)
frame $N$ to be similarly equipped with dextral unit vectors $\hat{n}_x,
\hat{n}_y, \hat{n}_z$, with $\hat{n}_z$ perpendicular to the ground plane
(downwards) and aligned with the local gravitational field. To orient the disc,
first align $C$ with $N$, then apply successive body fixed ZXY rotations:
yaw $q_1$, roll $q_2$, and spin $q_3$.  Denote by $A$ and $B$ the two frames
associated with the first two intermediate rotations in the sequence; we refer
to these as the instantaneous yaw and lean frames, respectively. The unit
vector, $\hat{b}_z$ fixed in the $B$ frame, is parallel to the disc plane and
aligned with the position vector from the disc center $C^*$ to the lowest point
of the disc $P^*$.

Traditionally, to locate the center of the disc $C^*$
relative to the inertial origin $N^*$, two coordinates are used to locate the
contact point in the ground plane; these two coordinates and the lean angle
and yaw angles determine the location of the center of the disc. The advantage of this
approach is that the choice of coordinates guarantees that the disc remain in
contact with the ground plane.

To illustrate how dependent coordinates must be
accounted for when linearizing the equations of motion, we purposefully locate
the center of the disc $C^*$ relative to the inertial origin $N^*$ with a
non-minimal choice of coordinates
\begin{equation*}
  \bar{r}^{N^*C^*} = q_4 \hat{n}_x + q_5 \hat{n}_y + q_6 \hat{n}_z
\end{equation*}
which results in the configuration constraint $\mathbf{f}_c$
\begin{equation}
  \label{rd:f_c}
  r\cos{q_2} + q_6 = 0
\end{equation}
which must be satisfied for the disc to remain in contact with the ground.
While this constraint is easily avoided by appropriate choice of coordinates,
it serves to illustrate considerations which must be made when analyzing
systems in which there is no obvious way to eliminate configuration constraints.

Again to demonstrate how velocity constraints must be accounted for when
obtaining linearized equations of motion, we define the velocity of $C^*$ in
$N$ with a non-minimal choice of speeds, defined in the body-fixed ($C$) frame.
The angular velocity of $C$ in $N$, however, is defined in the intermediate
frame $B$, which is the instantaneous lean frame.
\begin{align}
  \label{v_u}
  {^N}\bar{v}^{C^*} &= u_4 \hat{c}_x + u_5 \hat{c}_y + u_6 \hat{c}_z \\
  \label{w_u}
  {^N}\bar{\omega}^C &= u_1 \hat{b}_x + u_2 \hat{b}_y + u_3 \hat{b}_z
\end{align}
The unconstrained partial angular velocity and partial velocity of $C^*$,
respectively, are then
\begin{align}
  {^N}\bar{v}^{C^*}_u &= [0, 0, 0, \hat{c}_x, \hat{c}_y, \hat{c}_z] \\
  {^N}\bar{\omega}^C_u &= [\hat{b}_x, \hat{b}_y, \hat{b}_z, 0, 0, 0]
\end{align}
where we present the partial velocities as a $1\times6$ matrix whose entries
are populated with standard 3-vectors. Given these definitions, the velocity of
the lowest point of the disc $P^*$ is
\begin{align*}
    {^N}\bm{v}^{P^*} &= \bm{v}^{C^*} + \bm{\omega}^{C} \times r^{P^*/C^*} \\
                     &= u_4\hat{c}_x + u_5\hat{c}_y + u_6\hat{c}_z + r u_2
                        \hat{b}_x - r u_1 \hat{b}_y \\
                     &= (r u_2 \cos(q_3) + u_4) \hat{c}_x + (-r
                        u_1 + u_5)\hat{c}_y + r u_2 \sin(q_3) +
                        u6) \hat{c}_z
\end{align*}
where we have made use of the fact that $\bm{r}^{P^*/C^*} = r\bm{b}_z$.
Under the assumption of pure rolling this immediately yields three velocity
constraint equations which make up $\mathbf{f}_v = \mathbf{0}$
\begin{subequations}
\label{rd:f_v}
\begin{align}
    r u_2 \cos(q_3) + u_4 &= 0\\
            -r u_1 + u_5 &= 0\\
    r u_2 \sin(q_3) + u_6 &= 0
\end{align}
\end{subequations}
These equations are linear in the $u_i$ terms, nonlinear in the $q_i$ terms,
involve geometric system parameters (but not mass or inertial parameters), and
do not explicitly involve $\dot{q}_i$ terms. This structure will be taken
advantage of subsequently.

At this point, the reader familiar with the rolling disc problem is certainly
wondering why we have chosen such a diabolical set of coordinates and speeds.
In this simple example, there is an {\bf \textit{obvious and minimal}} choice of
coordinates and speeds ($q_6, u_4, u_5, u_6$, should be dependent; this choice
will be valid for all parameters and configurations; further, the derivation of
the equations of motion can be done without ever introducing these quantities),
for some systems this is not the case (e.g., Stewart's platform, bicycle). The
numeric values of system parameters and the configurations in which the system
will operate will determine which coordinates and speeds can be taken to be
dependent (and hence potentially eliminated by algebraic manipulations).
However, it may not be possible \textit{apriori} to determine easily which
coordinates and speeds should be taken as dependent; further, some systems may
operate in distinct enough regions of the configuration space that one
\textit{must} use different choices of dependent coordinates and/or speeds as
the system moves from one region to another. The purpose of our complex
choice of coordinates and speeds is to illustrate, in the context of a familiar
example, how constraints must be taken into account when linearizing equations
of motion. It is our hope that by doing this for a well known and relatively
simple example, readers can apply the same techniques to more complicated
systems where its use is more appropriate or necessary.
